---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
We start with the data as a zip file. We need to create a tempfile in R and use the `unz()`
function to access the CSV file in the zip:

```{r}
activityTable <- read.csv(unz("activity.zip", "activity.csv"))
str(activityTable)
```

Now we need to pre-process the data in a couple of different ways. First of all, the dates are factors. We want them as dates. Secondly, the intervals are integers, but they represent clock values (135 represents 1:35 am - you'll notice there is no 60 or 160 interval). We want to turn those into true integers - the number of minutes since the start of the day. We effect these two transformations like so:

```{r}
activityTable$date <- as.Date(activityTable$date)
# This one's a little more complicated - interpret it as a string, break into two, then
# back into an integer:
minutesSinceStartOfDay <- function(intervalAsClockTime) {
        intervalString <- toString(intervalAsClockTime)
        stringLength <- nchar(intervalString)
        if (stringLength <= 2) {
                # This is easy; no hour (hour = 0), so just return our input
                return(intervalAsClockTime)
        }
        minutes <- as.integer(substr(intervalString, stringLength - 1, stringLength))
        hours <- as.integer(substr(intervalString, 1, stringLength - 2))
        return(hours * 60 + minutes)
}


## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
