---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
We start with the data as a zip file. We need to use the `unz()` function to access the CSV file in the zip:

```{r}
activityTable <- read.csv(unz("activity.zip", "activity.csv"))
str(activityTable)
```

Now we need to pre-process the data in a couple of different ways. First of all, the dates are factors. We want
them as dates. Secondly, the intervals are integers, but they represent clock values (135 represents 1:35 am -
you'll notice there is no 60 or 160 interval). We want to turn those into true integers - the number of minutes
since the start of the day. We effect these two transformations like so:

```{r}
activityTable$date <- as.Date(activityTable$date)
# This one's a little more complicated - interpret it as a string, break into two, then
# back into an integer:
minutesSinceStartOfDay <- function(intervalAsClockTime) {
        intervalString <- toString(intervalAsClockTime)
        stringLength <- nchar(intervalString)
        if (stringLength <= 2) {
                # This is easy; no hour (hour = 0), so just return our input
                return(intervalAsClockTime)
        }
        minutes <- as.integer(substr(intervalString, stringLength - 1, stringLength))
        hours <- as.integer(substr(intervalString, 1, stringLength - 2))
        return(hours * 60 + minutes)
}
# For future use:
clockTimeForMinutes <- function(totalMinutes) {
        hours <- trunc(totalMinutes / 60)
        minutes <- totalMinutes %% 60
        sprintf("%02d:%02d", hours, minutes)
}
activityTable$intervalStartInMinutes <- sapply(activityTable$interval, minutesSinceStartOfDay)
activityTable <- subset(activityTable, select=-c(interval))
str(activityTable)
```
## What is mean total number of steps taken per day?
To calculate the total number of steps per day, we group by day and sum steps. We'll create a histogram and
output the mean and median.
```{r}
stepsPerDay <- aggregate(. ~ date, data=activityTable, FUN=sum)
meanStepsPerDay <- mean(stepsPerDay[, "steps"])
medianStepsPerDay <- as.integer(median(stepsPerDay[, "steps"]))
hist(stepsPerDay[, "steps"], 8, xlab="Steps per day", ylab="Number of days",
     main="Histogram of number of steps per day")
```

The mean steps per day is `r sprintf("%.1f", meanStepsPerDay)` and the median is `r medianStepsPerDay`.
These are very similar, so we can infer that the data is evenly divided around the mean (the histogram
supports this).


## What is the average daily activity pattern?
Find the mean of steps for each interval over all days, and the interval with the highest mean:
```{r}
meanStepsPerInterval <- aggregate(. ~ intervalStartInMinutes, data=activityTable, FUN=mean)

maxMeanStepsIndex <- which.max(meanStepsPerInterval$steps)
maxMeanStepsIntervalStart <- meanStepsPerInterval[maxMeanStepsIndex, "intervalStartInMinutes"]

plot(meanStepsPerInterval$intervalStartInMinutes, meanStepsPerInterval$steps, type="l",
     xlab="Minutes since start of day", ylab="Number of steps", main="Mean steps for each interval")
```

The interval with the most steps, on average, is `r maxMeanStepsIntervalStart` minutes from the start of the
day; in other words, `r clockTimeForMinutes(maxMeanStepsIntervalStart)`.


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
